---
title: "Student Loan Relief Report"
author: "Amy Yates"
bibliography: references.bib
csl: nature.csl
link-citations: true
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
library(tidyverse)
library(janitor)
library(skimr)
```

# Background

As the cost of higher education continues to grow, student debt in America rose to $1.6 trillion in 2022, impacting 45 million individuals and their families [@koeze_russell_2022]. In August of 2022, the Biden Administration announced a debt relief program [@student_aid_gov] of $10000 for borrowers below a certain income level and $20000 for borrowers that had received Pell Grants. This initiative is currently blocked by courts, and monthly payments are paused until August of 2023 if litigation does not resolve sooner.

This report will explore higher education data to try and answer the following questions:

1. What are the characteristics of borrowers (demographics, income level, institutions, post-graduate level)?
1. Under the proposed plan, which borrowers are most positively impacted?
1. Is the income level cap of $125,000 per individual/$250,000 per family backed by data?
1. Can we identify recommendations for changes to the plan?

Some of the data available that may be explored is listed below:

* https://nces.ed.gov/ipeds/datacenter/DataFiles.aspx?year=2001
* https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-11-26
* https://tuitiontracker.org/
* https://data.ed.gov/dataset/federal-student-loan-portfolio/resources

# Pell Grant Data

*Source: https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-08-30*

Pell Grants are provided to students that meet certain qualifications for financial need. These are grants, not loans, and do not require repayment. These grants typically make only a small dent in the overall cost, so most students accepting Pell Grants also take out loans. Qualification for these grants serve as an indicator that these students lack the generational wealth that many college-bound individuals enjoy and that loan forgiveness would make a greater impact for them.

The average grant amount can vary widely by institution and by state. Since the federal grant program does not consider institutional cost, it is expected that income level of applicants is the main contributor to the variance. However, different institutions may offer better support for accessing these grants.

Below is a summary of the 2017 averages per state, ordered from the largest total per recipient down.

```{r}
pell <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-08-30/pell.csv') %>% clean_names()
total_per_recipient <- pell %>% 
  filter(year == 2017) %>% 
  group_by(state, year) %>% 
  summarise(award_total = sum(award), recipient_total = sum(recipient), total_per_recipient = award_total/recipient_total) %>% 
  ungroup() %>%  
  arrange(desc(total_per_recipient))
total_per_recipient
```
Six of the top ten states are in fact territories of the U.S. with Puerto Rico topping the list. This strengthens the supposition that populations with lower income levels are the primary contributor to larger individual grants. The top ten also includes New York and New Jersey, however, states that were in the top 1/3 of incomes in that same year.

To understand why that might be, let's take a look at the New York grants at an institutional level, ordering from high to low.

```{r}
inner_join(pell %>% 
    filter(state == 'NY') %>% 
    filter(year==2017) %>% 
    mutate(institution_per_recipient = award/recipient) %>% 
    select(state, name, institution_per_recipient), 
  total_per_recipient %>% select(state, total_per_recipient)
  ) %>% arrange(desc(institution_per_recipient))
```
The maximum Pell Grant amount that could be given to individuals in 2017 was $5920, yet the top 20 schools have averages far greater than that number. This could be reporting issues, but it is also possible to receive more than 100% of the maximum grant in a given year as long as the average over the years balances out to less than the maximum. This may add support for our other hypothesis - that some institutions are simply better at championing their students and maximizing the benefit.

# References
<div id="refs"></div>